Human tissue specific markers

### 2017-01-03 manuscript/manuscript.Rnw
fix cache and data loading

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{txfonts}
\usepackage{fancyref}
\usepackage{acro}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=numeric-comp]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\usepackage[capitalize]{cleveref}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
\pagestyle{fancy}

\newcommand{\EV}{\EV}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
#opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -1, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("proto")
library("xtable")
library("caret")
library("pROC")
library("grid")

@@

<<load.data,echo=FALSE,results="hide",warning=FALSE,cache.extra=tools::md5sum(paste0("../tissue_specific_expression/",c("tissue_specific_markers","interesting_gene_reads_wide","combined_star_logs","svm_test_results","svm_test_results_reduced","housekeeping_genes_superset")))>>=
for (a in paste0("../tissue_specific_expression/",
                 c("tissue_specific_markers",
                   "interesting_gene_reads_wide",
                   "combined_star_logs",
                   "svm_test_results",
                   "svm_test_results_reduced",
                   "housekeeping_genes_superset"
                   ))) {
    load(a)
}
tissue.specific.genes <- data.table(tissue.specific.genes)
setkey(tissue.specific.genes,"gene")
@

@@

\EV are small (40-1000nm) compartments that can be found in
blood, saliva, and urine. \EV originate from the plasma membrane
or multivesicular endosome of many cell types and contain proteins,
lipids, mRNA, miRNA, and many other non-coding RNAs. Because \EV
contain mRNA from the cell of origin and can be obtained without
invasive procedures, they are a promising remote sensor of the
transcriptome of tissues that would otherwise be inaccessible.

@@

Reads were aligned to \textit{Homo sapiens} reference genome GRCh38
with Ensembl annotation release 80 using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and called using
Tophat 2 v2.0.10
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}. The average
percentage of uniquely mapped reads was %
\Sexpr{mean(as.numeric(gsub("%","",star.logs[field=="Uniquely mapped reads %",value])))}\%
and the average total mapping percentage was %
\Sexpr{mean(as.numeric(gsub("%","",star.logs[field=="Uniquely mapped reads %",value]))+as.numeric(gsub("%","",star.logs[field=="% of reads mapped to multiple loci",value])))}\%.
% 
For samples where the reads were not publicly available (GTEx and
some Roadmap Epigenomics samples), the RPKM were taken from publicly
available result files (see code for details).

@@

The population of RNAs found in \EV are distinct from the cytosolic
pool, and are likely selectively sorted into \EV~%
\cite{Villarroya-Beltri.ea2014:Sortingitoutregulation,Hoen.ea2012:DeepsequencingofRNA,Janas.ea2015:MechanismsofRNAloading,Balkom.ea2015:Quantitativeandqualitativeanalysis}.
Specifically, the motives
GGAG~\cite{Villarroya-Beltri.ea2013:SumoylatedhnRNPA2B1controlssorting},
ACCAGCCU, CAGUGAGC, and
UAAUCCCA~\cite{Batagov.ea2013:Exosomessecretedbyhuman} are enriched in
\EV. Owing to the limited number of tissue-specific \ac{EV} RNA-seq
results, we have biased the reads from existing tissue-specific
RNA-seq results to preferentially include RNAs which contain motifs
found in exosomes and also match the lower number of unique RNAs which
are found in individual exosomes.

@@

The \ac{SVM} was generated using caret
\cite{Kuhn2008:Buildingpredictivemodelsinr} and svmRadial
\cite{Karatzoglou2004:kernlab} with 10 cycles of 10 repeats and a tune
length of 8 with a training set representing $\approx 75\%$ of the
samples ($\approx 2450$ samples total, including $\approx 35$ uterine
or $\approx 60$ placenta) in each of the groups chosen at random, with
the remaining samples making up a testing set ($\approx 820$ samples).

@@

\begin{figure}
<<tissue.specificity,shape.figure,fig.width=8,fig.height=8,echo=FALSE>>=
invisible(tissue.specific.genes[,type:="not specific"])
invisible(tissue.specific.genes[,housekeeping:="not housekeeping"])
invisible(tissue.specific.genes[possibly.specific==TRUE,type:="tissue specific"])
invisible(tissue.specific.genes[tracking_id %in% housekeeping.genes.superset[,tracking_id] &
                                 possibly.specific==FALSE,housekeeping:="housekeeping"])
invisible(tissue.specific.genes[tracking_id %in% housekeeping.genes.superset[,tracking_id] &
                          possibly.specific==TRUE,housekeeping:="housekeeping and specific"])

@@

legend("bottomright",       legend=paste0(capwords(a)," ",c("1 Read","10 Reads","1000 Reads", "5000 Reads")),
       col=c("brown","green","#002058", "#E87722"),
       lwd=2)
}
grid.text(c("A","B","C","D"),
          x=unit(c(0,0.5,0,0.5),"npc"),
          y=unit(c(1,1,0.5,0.5),"npc"),
          just=c("left","top"),
          gp=gpar(fontsize=32)
          )

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{txfonts}
\usepackage{fancyref}
\usepackage{acro}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\usepackage[capitalize]{cleveref}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -1, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("proto")
library("xtable")
library("caret")
library("pROC")
library("grid")

@@

<<load.data,echo=FALSE,results="hide",warning=FALSE,cache.extra=file.info(paste0("../tissue_specific_expression/",c("tissue_specific_markers","interesting_gene_reads_wide","combined_star_logs","svm_test_results","svm_test_results_reduced","housekeeping_genes_superset")))[,"mtime"]>>=
for (a in paste0("../tissue_specific_expression/",
                 c("tissue_specific_markers",
                   "interesting_gene_reads_wide",
                   "combined_star_logs",
                   "svm_test_results",
                   "svm_test_results_reduced",
                   "housekeeping_genes_superset"
                   ))) {
    load(a)
}
tissue.specific.genes <- data.table(tissue.specific.genes)
setkey(tissue.specific.genes,"gene")
@

@@

\Ac{EV} are small (40-1000nm) compartments that can be found in
blood, saliva, and urine. \Ac{EV} originate from the plasma membrane
or multivesicular endosome of many cell types and contain proteins,
lipids, mRNA, miRNA, and many other non-coding RNAs. Because \ac{EV}
contain mRNA from the cell of origin and can be obtained without
invasive procedures, they are a promising remote sensor of the
transcriptome of tissues that would otherwise be inaccessible.

@@

Reads were aligned to \textit{Homo sapiens} reference genome GRCh38
with Ensembl annotation release 80 using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and called using
Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}. The average
percentage of uniquely mapped reads was %
\Sexpr{mean(as.numeric(gsub("%","",star.logs[field=="Uniquely mapped reads %",value])))}\%
and the average total mapping percentage was %
\Sexpr{mean(as.numeric(gsub("%","",star.logs[field=="Uniquely mapped reads %",value]))+as.numeric(gsub("%","",star.logs[field=="% of reads mapped to multiple loci",value])))}\%.
% 
For samples where the reads were not publicly available (GTEx and
some Roadmap Epigenomics samples), the RPKM were taken from publicly
available result files (see code for details).

@@

The population of RNAs found in \ac{EV} are distinct from the cytosolic
pool, and are likely selectively sorted into \ac{EV}~%
\cite{Villarroya-Beltri.ea2014:Sortingitoutregulation,Hoen.ea2012:DeepsequencingofRNA,Janas.ea2015:MechanismsofRNAloading,Balkom.ea2015:Quantitativeandqualitativeanalysis}.
Specifically, the motives
GGAG~\cite{Villarroya-Beltri.ea2013:SumoylatedhnRNPA2B1controlssorting},
ACCAGCCU, CAGUGAGC, and
UAAUCCCA~\cite{Batagov.ea2013:Exosomessecretedbyhuman} are enriched in
\ac{EV}. Owing to the limited number of tissue-specific \ac{EV} RNA-seq
results, we have biased the reads from existing tissue-specific
RNA-seq results to preferentially include RNAs which contain motifs
found in exosomes and also match the lower number of unique RNAs which
are found in individual exosomes.

@@

The classification of samples into reads of placental origin vs
non-placental origin or uterine origin vs non-uterine origin was
performed using a \ac{SVM}. The \ac{SVM} was generated using caret
\cite{Kuhn2008:Buildingpredictivemodelsinr} and svmRadial
\cite{Karatzoglou2004:kernlab} with 10 cycles of 10 repeats and a tune
length of 8 with a training set representing $\approx 75\%$ of the
samples ($\approx 2450$ samples total, including $\approx 35$ uterine
or $\approx 60$ placenta) in each of the groups chosen at random, with
the remaining samples making up a testing set ($\approx 820$ samples).

@@

\begin{figure}
<<tissue.specificity.shape.figure,fig.width=8,fig.height=8,echo=FALSE>>=
invisible(tissue.specific.genes[,type:="not specific"])
invisible(tissue.specific.genes[,housekeeping:="not housekeeping"])
invisible(tissue.specific.genes[possibly.specific==TRUE,type:="tissue specific"])
invisible(tissue.specific.genes[tracking_id %in% housekeeping.genes.superset[,tracking_id] &
                                 possibly.specific==FALSE,housekeeping:="housekeeping"])
invisible(tissue.specific.genes[tracking_id %in% housekeeping.genes.superset[,tracking_id] &
                          possibly.specific==TRUE,housekeeping:="housekeeping and specific"])

@@

legend("bottomright",       legend=paste0(capwords(a)," ",c("1 Read","10 Reads","1000 Reads", "5000 Reads")),
       col=c("brown","green","#002058", "#E87722"),
       lwd=2)
}
### grid.text(c("A","B","C","D"),
###           x=unit(c(0,0.5,0,0.5),"npc"),
###           y=unit(c(1,1,0.5,0.5),"npc"),
###           just=c("left","top"),
###           gp=gpar(fontsize=32)
###           )
